Predicting the Outcome of Patient-Provider Communication Sequences using Recurrent Neural Networks and Probabilistic Models

### 2018-01-10 paper-amia-summit-2018/amia_summit_2018.tex
camera ready version

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered sequences of observations generated by a molecular, physiological or psychological process to make predictions about the outcome of that process arises in many domains of clinical informatics. In this paper, we focus on predicting the outcome of patient-provider communication sequences in the context of clinical dialog. Specifically, we consider prediction of the motivational interview success (i.e. eliciting a particular type of patient behavioral response) based on an observed sequence of coded patient-provider communication exchanges as a sequence classification problem. We propose two solutions to this problem, one that is based on Recurrent Neural Networks (RNNs) and another that is based on Markov Chain (MC) and Hidden Markov Model (HMM), and compare the accuracy of these solutions using communication sequences annotated with behavior codes from the real-life motivational interviews. Our experiments indicate that the deep learning-based approach is significantly more accurate than the approach based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM, respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling). These results indicate that the proposed method can be used for real-time monitoring of clinical interviews progression and more efficient identification of effective provider communication strategies, which can significantly decrease the effort required to develop behavioral interventions and increases their effectiveness.}

\section*{Introduction}
Temporally ordered sequences of discrete or continuous observations generated by molecular, psychological or psychological process(es) arise in many different areas of biology and medicine (e.g., DNA base-pairs, protein sequences, ECG measurements, laboratory results, diagnostic codes, utterances  in clinical dialog). Classification (or categorization) is a type of analysis of those sequences that has a broad range of important practical applications, from protein function \cite{yakhnenko2005discriminatively} or structure \cite{deshpande2002evaluation} prediction to detecting individuals with a heart disease \cite{wei2006semi}. Taking into account both the entire set of observations in a sequence, as well as temporal order and potential dependencies between observations, makes sequence classification a more challenging task than classification of independent observations. Predicting the outcome of those sequences (e.g. physiological or behavioral response) can also be viewed as a sequence classification problem.

In general, sequence classification methods fall into one of the three major classes: feature-based, distance-based and model-based. Feature-based methods transform a sequence into a feature vector and apply a standard supervised machine learning method, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at a classification decision. The methods in this class have had limited success, since traditional feature representation methods cannot easily account for the order of and dependencies between observations in a
sequence.

@@

In this paper, we address the problem of predicting the outcome of coded patient-provider communication sequences in the context of clinical dialog. Specifically, we focus on predicting the success (i.e. eliciting a particular type of patient behavioral response) of motivational interviews with obese adolescents and their caregivers based on an observed sequence of coded patient-provider communication exchanges during those interviews. Childhood obesity is a serious public health concern in the United States. Recent estimates indicate that approximately one-third (31.8\%) of U.S. children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents, who are obese, are likely to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, collectively referred to as ``change talk'' (CHT), consistently predict the actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of providers to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual examination and analysis of MI interview transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective behavioral interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

Our goal is to compare the accuracy of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for the task of predicting the success of clinical interviews (i.e. eliciting a particular type of patient behavioral response, such as CHT) at any point during a clinical interview based on a sequence of coded previous patient-provider communication exchanges in the same interview, which we consider as a sequence classification problem. This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook \cite{carcone2013provider}. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability of state-of-the-art methods for sequence modeling to the analysis of patient-provider communication (PPC) exchanges, in general, and predicting the desired patient behavioral response in the context of motivational interviews, in particular.

@@

The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which consist of a total of 50,239 segmented and annotated utterances. Each transcript consists of an MI interview session, which typically involves a counselor, an adolescent and a caregiver. The utterances were annotated based on the MYSCOPE codebook \cite{carcone2013provider}, in which the behavior codes are grouped into the patient (adolescent and caregiver) codes and the counselor codes. Annotated utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences are the ones, which resulted in positive change talk (CHT+) or commitment language (CML+) statements by an adolescent or a caregiver, while unsuccessful sequences are the ones, which resulted in negative change talk (CHT-) or commitment language (CML-), or the ones, in which no change talk or commitment language statements were made.

A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. In this example, $SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$ is a successful sequence, in which a counselor starts with an open-ended question and ultimately is able to elicit a positive change talk statement. As follows from this example, similar utterances, such as ``Yeah'' and ``Yes'', can be assigned different behavior codes (CHT+ and HUPW, respectively), depending on the context.\\

@@

Since severely imbalanced datasets often distort the true performance of a classification method relative to a simple ``majority vote'' baseline (e.g. simply classifying every communication sequence as successful would result in 82.15\% accuracy on our dataset). Therefore, it is important to
properly address class imbalance. We evaluated the performance of probabilistic and deep learning methods using both under-sampling and over-sampling for balancing the number of samples in different classes. Synthetic Minority Over Sampling Technique (SMOTE) \cite{chawla2002smote} is a widely used oversampling method for imbalanced datasets, in which new synthetic examples are generated for minority classes. Specifically, we generated synthetic examples at the borderline between the majority and minority classes \cite{nguyen2011borderline}. On the other hand, the under-sampling method reduces the number of samples in majority class by replacing the clusters of samples identified by the $k$-means clustering algorithm with the cluster centroids.

Two MC and HMM models were trained, one model was trained on successful sequences, while another was trained on unsuccessful sequences.

\subsection*{\textit{Prediction method}}
Generally, a sequence can be viewed as a temporally ordered set of events. In this study, an event is a behavior code that also has a symbolic representation, such as $LUP+$ (low uptake, positive),
$OQECHT+$ (open-ended question, elicit change talk positive), etc.  Given a sequence of behavior codes $S_i = \{c_1, c_2,...,c_n\}$ representing patient-provider communication exchanges during some
part of a motivational interview, the task of predicting interview success can be viewed as sequence classification. Given a set of class labels $L = \{l_1, l_2,...,l_m\}$ (in our case, the labels are
``successful'' and ``unsuccessful'' motivational interview), a sequence classifier $C$ learns a function $S_i \to l_i, l_i \in L$ that maps a sequence $S_i$ into a class label $l_i \in L$.

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first and second-order Markov Chain and
Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classify each test sequence based on the maximum
likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence
modeling \cite{mutsam2016maximum, won2004training}. However, the latest advances in deep learning indicate that RNNs provide better results than conventional machine learning methods for the task of
sequence classification. Therefore, we employed two special variants of RNN in our experiments: Long Short-Term Memory (LSTM) \cite{hochreiter1997long} and Gated Recurrent Unit (GRU)
\cite{chung2014empirical}.

\textbf {Markov Chain (MC)} is a probabilistic model that predicts the probability of next state based on its current state but not on its past states (Markov property). For the sequential analysis,
we built two Markov models $M$ and $\overline{M}$, summarizing counselor strategies and patient responses, in the cases of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A
Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
\begin{itemize}
\item $V = \{CML+, CHT+, CHT-, AMB-, LUP+, LUP-, HUPW, OQO, CQTBN, CQECHT+,...\}$ is a set of vertices, consisting of adolescent, caregiver and counselor MI behavior codes;
\item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:
\begin{equation}
P_M(c_j|c_i) = \frac{n_{c_i,c_j}}{n_{c_i}}
\end{equation}
\end{itemize}
where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ has been observed in the training data. Given a Markov model $M$ (such that $S\subseteq V$), the probability that a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ has been generated from a Markov model $M$ is:
\begin{equation}
P_M(S) = \prod_{i=2}^N p_M(c_i|c_1,\dots,c_{i-1})=\prod_{i=2}^N p_M(c_i|c_{i-1})
\end{equation}
In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point as:
\begin{equation}
p(S\rightarrow successful) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}
If $p(S\rightarrow successful) > 0 $, a communication sequence is predicted to result in positive change talk or commitment language. Otherwise, it would be classified as a negative change talk or
commitment language.

The above model also referred as first-order MC, since it only considers immediately preceding behavior code, when computing the state transition probabilities. In our experiment, we also considered
second-order Markov model that computes transition probabilities based on previous two states.

\textbf {Hidden Markov Model (HMM)} is another probabilistic model used for modeling processes varying in time. HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. Mathematically, HMM can be defined as the following: $\lambda = (A, B, \pi)$.
\begin{itemize}
\item A is an $N\times N$ state transition probability distribution matrix $A = \{a_{ij}\}$
\item B is an $N\times M$ matrix $B = \{b_j(k)\}$ with observation symbol probability distribution for each state 
\item $\pi$ is the initial state distribution vector $\pi = \{\pi_i\}$
\end{itemize}
Hence, $N$ is a number of hidden states in the model and $M$ is a number of distinct observations symbols per state, i.e. the discrete vocabulary size. The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols (emission probabilities) for each state. The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}.

\textbf {Embeddings}: We took the inspiration for the representation of behavior codes from the idea of word embeddings\cite{bengio2003neural}. Word embedding is a representation of words in
low-dimensional space by vectors, which contain the features of the words. In our study, we employed embedding in place of one-hot vectors for representation of behavior codes as input to LSTM and
GRU, since one-hot vectors are high-dimensional and sparse. Moreover, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by
retaining their internal relations. The embedding was learned during the training of the classification model. Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embedding in 44-dimensional
vector space visualized by t-SNE \cite{maaten2008visualizing}. It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc. formed a cluster in the left
part of the figure. The nearest neighbors of CQECML+ are highlighted by their color intensity (i.e. OQECML+ being more purple means closer to CQECML+). The right part of the figure demonstrates
another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc. It is interesting that the behaviors intended
to elicit CHT+/CML+ group together whereas the ones intended to elicit CHT-/CML- also group together and are located on opposite ends of the space.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
    \caption{\textbf{T-SNE diagram of behavior codes drawn after 4000 steps with perplexity 30}.}
    \label{fig:code_embedding}
\end{figure}

\textbf {Recurrent Neural Networks (RNN)} are a type of Neural Networks, which can send feedback from the current hidden state to the hidden state of the next time step. The ability of RNN to capture
long-term dependencies for predicting the future events is its main advantage over MC and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior
observed at some point in the interview is very informative for the future behaviors, that will be observed later, but not immediately. However, it was observed that RNN fails to capture long-term
dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short
Term Memory networks or simply LSTM. There are several variants of LSTM model, among which the most notable one is the Gated Recurrent Unit\cite{cho2014properties} (GRU). GRUs are
simpler than LSTM units and have been shown to experimentally outperform other models\cite{cho2014properties}. GRU is formally defined as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\label{eq:resetgru}
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\label{eq:candidategru}
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
As these equations, $\sigma$ stands for an element-wise application of the sigmoid function and $\odot$ is an element-wise product. The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W$ and $U$ represent weights and $b$ represents biases. The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}).
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/rnn_small.eps}
    \caption{\textbf{Proposed RNN model with target replication (TR)}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden layer of 15 LSTM or GRU nodes and embedding of MYSCOPE codes in an observed
sequence of communication behaviors as input. Then, we applied a softmax layer at each time-step for predicting the label of the sequence. Since the sequence label was predicted at each time step, the proposed model is also known as Recurrent Neural Network with Target Replication (RNN-TR). As can be seen from
Eq.~\ref{eq:loss}, the total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where $T$ is the total number of elements in a sequence,
$\bar y^{(t)}$ is the output at step $t$, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Experimentally, it was determined that
best performance achieved when we use $\alpha=0.5$. Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which
were determined by the validation set. We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges
to optimal after 100 epochs and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

>>>>

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered sequences of observations generated by molecular, physiological or psychological processes to make predictions about the outcome of these processes arises in many domains of clinical informatics. In this paper, we focus on predicting the outcome of patient-provider communication sequences in the context of clinical dialog. Specifically, we consider prediction of the motivational interview success (i.e. eliciting a particular type of patient behavioral response) based on an observed sequence of coded patient-provider communication exchanges as a sequence classification problem. We propose two solutions to this problem, one that is based on Recurrent Neural Networks (RNNs) and another that is based on Markov Chain (MC) and Hidden Markov Model (HMM), and compare the accuracy of these solutions using communication sequences annotated with behavior codes from the real-life motivational interviews. Our experiments indicate that the deep learning-based approach is significantly more accurate than the approach based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM, respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling). These results indicate that the proposed method can be used for real-time monitoring of progression of clinical interviews and more efficient identification of effective provider communication strategies, which in turn can significantly decrease the effort required to develop behavioral interventions and increase their effectiveness.}

\section*{Introduction}
Temporally ordered sequences of discrete or continuous observations generated by molecular, psychological or psychological process(es) arise in many different areas of biology and medicine (e.g., DNA base-pairs, protein sequences, ECG measurements, laboratory results, diagnostic codes, utterances  in clinical dialog). Classification (or categorization) is a type of analysis of those sequences that has a broad range of important practical applications, from protein function \cite{yakhnenko2005discriminatively} or structure \cite{deshpande2002evaluation} prediction to detecting individuals with a heart disease \cite{wei2006semi}. Taking into account both the entire set of observations in a sequence, as well as the temporal order and potential dependencies between observations, makes sequence classification a more challenging task than classification of independent observations. Predicting the outcome of those sequences (e.g. physiological or behavioral response) can also be viewed as a sequence classification problem.

In general, sequence classification methods fall into one of three major classes: feature-based, distance-based and model-based. Feature-based methods transform a sequence into a feature vector and apply a standard supervised machine learning method, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at classification decision. The methods in this class have had limited success, since traditional feature representation methods cannot easily account for the order of and dependencies between observations in a
sequence.

@@

In this paper, we address the problem of predicting the outcome of coded patient-provider communication (PPC) sequences in the context of clinical dialog. Specifically, we focus on predicting the success (i.e. eliciting a particular type of patient behavioral response) of motivational interviews with obese adolescents and their caregivers based on an observed sequence of coded PPC exchanges during those interviews. Childhood obesity is a serious public health concern in the United States. Recent estimates indicate that approximately one-third (31.8\%) of U.S. children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents, who are obese, are likely to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, collectively referred to as ``change talk'' (CHT), consistently predict the actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of providers to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual examination and analysis of MI interview transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective behavioral interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

Our goal is to compare the accuracy of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for the task of predicting the success of clinical interviews (i.e. eliciting a particular type of patient behavioral response, such as CHT) at any point during a clinical interview based on a sequence of coded previous PPC exchanges in the same interview. This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook \cite{carcone2013provider}. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability of state-of-the-art methods for sequence modeling to the analysis of PPC exchanges, in general, and predicting the desired patient behavioral response in the context of motivational interviews, in particular.

@@

The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which consist of a total of 50,239 segmented and annotated utterances. Each transcript corresponds to an MI interview session, which typically involves a counselor, an adolescent and a caregiver. The utterances were annotated based on the MYSCOPE codebook \cite{carcone2013provider}, in which the behavior codes are grouped into the patient (adolescent and caregiver) codes and the counselor codes. Annotated utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences are the ones, which resulted in positive change talk (CHT+) or commitment language (CML+) statements by an adolescent or a caregiver, while unsuccessful sequences are the ones, which resulted in negative change talk (CHT-) or commitment language (CML-), or the ones, in which no change talk or commitment language statements were made.

A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. In this example, $SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$ is a successful sequence, in which a counselor starts with an open-ended question and ultimately is able to elicit a positive change talk statement. As follows from this example, similar utterances, such as ``Yeah'' and ``Yes'', can be assigned different behavior codes (CHT+ and HUPW), depending on the context.\\

@@

Since severely imbalanced datasets often distort the true performance of a classification method relative to a simple ``majority vote'' baseline (e.g. simply classifying every communication sequence as successful would result in 82.15\% accuracy on our dataset), it is important to properly address class imbalance. We evaluated the performance of probabilistic and deep learning methods using both under-sampling and over-sampling for balancing the number of samples in different classes. Synthetic Minority Over Sampling Technique (SMOTE) \cite{chawla2002smote} is a widely used oversampling method for imbalanced datasets, in which new synthetic examples are generated for minority classes. Specifically, we generated synthetic examples at the borderline between the majority and minority classes \cite{nguyen2011borderline}. On the other hand, the under-sampling method reduces the number of samples in majority class by replacing the clusters of samples identified by the $k$-means clustering algorithm with the cluster centroids.

%Two MC and HMM models were trained, one model was trained on successful sequences, while another was trained on %unsuccessful sequences.

\subsection*{\textit{Sequence classification methods}}
In general, a sequence can be viewed as a temporally ordered set of observations. In this study, an observation corresponds to a behavior code, which has a symbolic representation, such as $LUP+$ (low uptake, positive),
$OQECHT+$ (open-ended question, elicit change talk positive), etc.  Given a sequence of behavior codes $S_i = \{c_1, c_2,...,c_n\}$ representing PPC exchanges during some part of a motivational interview, the task of predicting interview success can be considered as sequence classification. Given a set of class labels $L = \{l_1, l_2,...,l_m\}$ (in our case, the labels are
``successful'' and ``unsuccessful'' motivational interview), a sequence classifier $C$ learns a function $S_i \to l_i, l_i \in L$ that maps a sequence $S_i$ into a class label $l_i \in L$.

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first and second-order Markov Chain and
Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classify each test sequence based on the maximum
likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence
modeling \cite{mutsam2016maximum, won2004training}. However, the latest advances in deep learning suggest that RNNs may provide better results than conventional machine learning methods for the task of
sequence classification. To verify this hypothesis, we employed two state-of-the-art variants of RNN in our experiments: Long Short-Term Memory (LSTM) \cite{hochreiter1997long} and Gated Recurrent Unit (GRU) \cite{chung2014empirical}.

\textbf {Markov Chain (MC)} is a probabilistic model that conditions each observation in a sequence only on preceding observation and not on any other past observation. First, we estimated two Markov models $M$ and $\overline{M}$, summarizing counselor strategies and patient responses, in the cases of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A
Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
\begin{itemize}
\item $V = \{CML+, CHT+, CHT-, AMB-, LUP+, LUP-, HUPW, OQO, CQTBN, CQECHT+,...\}$ is a set of vertices, consisting of adolescent, caregiver and counselor MI behavior codes;
\item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on the maximum likelihood estimation:
\begin{equation}
P_M(c_j|c_i) = \frac{n_{c_i,c_j}}{n_{c_i}}
\end{equation}
\end{itemize}
where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the number of times the code $c_i$ have been observed in the training data, respectively. Given a Markov model $M$ (such that $S\subseteq V$), the probability that a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ has been generated from a Markov model $M$ is:
\begin{equation}
P_M(S) = \prod_{i=2}^N p_M(c_i|c_1,\dots,c_{i-1})=\prod_{i=2}^N p_M(c_i|c_{i-1})
\end{equation}
In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point as:
\begin{equation}
p(S\rightarrow successful) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}
If $p(S\rightarrow successful) > 0 $, a communication sequence is predicted to be successful (i.e. result in positive change talk or commitment language). Otherwise, it is predicted to be unsuccessful.

The above model is also referred as first-order MC, since it only considers immediately preceding behavior code, when computing the state transition probabilities. In our experiment, we also considered second-order Markov model, which conditions each observation on the preceding two observations.

\textbf {Hidden Markov Model (HMM)} is another probabilistic model used for modeling processes varying in time. HMMs are widely used for sequence analysis because of their ability to identify hidden states, corresponding to clusters of observations. Mathematically, HMM can be defined as $\lambda = (A, B, \pi)$, where:
\begin{itemize}
\item A is an $N\times N$ state transition probability distribution matrix $A = \{a_{ij}\}$
\item B is an $N\times M$ matrix $B = \{b_j(k)\}$ with observation symbol probability distribution for each state 
\item $\pi$ is the initial state distribution vector $\pi = \{\pi_i\}$
\end{itemize}
Hence, $N$ is a number of hidden states in the model and $M$ is a number of distinct observations per hidden state, i.e. the discrete vocabulary size. The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter. HMM deduces a sequence of hidden states that best explains the observations along with the state transition probabilities and the distributions of observations (emission probabilities) per each hidden state. The Baum-Welch algorithm \cite{rabiner1989tutorial} is used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm \cite{rabiner1989tutorial} is used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of success for an interview can be estimated using Eq.~\ref{eq:class} as well.

\textbf {Behavior code embeddings}. Representation of behavior codes was inspired by the recent success of word embeddings\cite{bengio2003neural}. Embedding is a representation of an object in low-dimensional space using a real-valued vector. In our study, embeddings of behavior codes were obtained as a by-product of training LSTM and GRU after feeding one-hot vectors as representation of behavior codes as input to these RNNs. Behavior code embeddings have the property of representing similar codes with the vectors that are close to each other in low-dimensional space. Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embeddings visualized in 2-dimensional space by t-SNE \cite{maaten2008visualizing}. It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc. formed a cluster in the left part of Figure~\ref{fig:code_embedding}. The nearest neighbors of CQECML+ are highlighted by different color intensity (i.e. OQECML+ being more purple indicates that it is more similar to CQECML+). The right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc. It is interesting that the behaviors intended to elicit CHT+/CML+ group together, whereas the ones intended to elicit CHT-/CML- also group together and are located in the opposite regions of semantic space.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
    \caption{\textbf{2-D representation of behavior code embeddings}.}
    \label{fig:code_embedding}
\end{figure}

\textbf {Recurrent Neural Networks (RNN)} are a class of neural networks that have an internal memory, which makes them particularly suitable for processing sequences of observations. The ability of RNNs to capture long-term dependencies and remember past observations for predicting future observations is their main advantage over MCs and HMMs. These features are very useful in analysis of motivational interviews, in which any behavior observed at a particular point in the interview may be indicative of other behaviors that are observed later. In order to mitigate the vanishing gradient problem of earlier versions of RNN \cite{bengio1993problem}, Hochreiter et al.\cite{hochreiter1997long} proposed Long Short Term Memory networks (LSTM). There are several variants of LSTM model, among which the most notable one is the Gated Recurrent Unit\cite{cho2014properties} (GRU). GRUs are simpler than LSTMs and have been shown to be effective for a variety of Natural Language Processing tasks \cite{cho2014properties}. GRU is formally defined as follows:
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\label{eq:resetgru}
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\label{eq:candidategru}
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
In Eq.~\ref{eq:firstgru}-\ref{eq:lastgru}, $\sigma$ corresponds to sigmoid function and $\odot$ designates an element-wise product. The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W_z$, $W_r$, $W_h$, $U_z$, $U_r$, $U_h$ are the weight matrices and $b_z$, $b_h$ and $b_r$ are bias vectors. The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}).

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/rnn_small.eps}
    \caption{\textbf{Proposed RNN model with target replication (TR)}.}
    \label{fig:rnn-model}
\end{figure}

The RNN architecture employed for sequence classification is shown in Figure~\ref{fig:rnn-model}. As can be seen from Figure~\ref{fig:rnn-model}, softmax is used at each time step to predict the class of a sequence observed so far. Since the sequence label is predicted at each observation, the proposed architecture is referred to as Recurrent Neural Network with Target Replication (TR). It is trained by minimizing the following hybrid loss function:

\begin{equation}
\tilde{\mathcal{L}} = \alpha \cdot \frac{1}{T}\sum_{t=1}^T \mathcal{L}(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot \mathcal{L}(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}

As follows from Eq.~\ref{eq:loss}, the total loss $\tilde{\mathcal{L}}$ is a convex combination of the final loss $\mathcal{L}(\bar y^{(T)},y^{(T)})$ and the average loss over all observations in a sequence, where $T$ is the total number of observations, $\bar y^{(t)}$ is the output at step $t$, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter controlling the relative importance of each loss type. We experimentally determined that the best performance is achieved when $\alpha=0.5$. Our model also contains several other hyperparameters, such as the number of embedding dimensions, the number of hidden units, learning rate, batch size, etc., which were optimized on the validation set. We implemented our models in Tensorflow with Adam optimizer as well as early stopping based on the validation loss and observed that our model converges after 100 epochs.
